Project 1 for Complex Adaptive Systems

### 2018-04-03 paper.tex
Late push

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
We investigate information flow between higher and lower levels flow of information in models of population growth. 
 We attempt to replicate the results found in \cite{Walker} and DO OTHER STUFF. 
\end{abstract}

@@

\noindent For individual populations, we assume birth rate, death rate, and  carrying capacity are all fixed.  Define $k$ to be the carrying capacity of a population and $n_t$ to be size of the population at time $t$. Then the size of the population at generation $t$ is modeled by 
\begin{eqnarray}
n_{t+1} = (\textnormal{birthrate} - \textnormal{deathrate}) (k n_t - n_t^2) / k 
\end{eqnarray}
If we define the population size relative to the carrying capacity as  $x_t = n_t / k$ and define the reproductive fitness  $r = birthrate - deathrate$, then we  get the following equation for $x_t$: 
\begin{eqnarray}
x(t+1) = x_{t+1} = r  x_t (1 - x_t), \ \ \ \ \  t=0, 1, 2, \ldots . 
\label{logistic}
\end{eqnarray}
This is the logistic map as defined in \cite{Mitchell}.  In what follows, we will refer to $x_t$ as the population size, with the understanding that it will always be the ratio of the actual population size to the carrying capacity

In \cite{Walker}, the authors investigate information flow between higher and lower levels of organization, between individuals of a population and the population considered as a whole. 
They study a model that consists of a population of individual populations and examine the direction of the causal information transfer, where 
"non-trivial collective behavior is associated with the degree to which local elements receive information from the global network." \cite{Walker}

Their model, a lattice of  globally coupled logistic maps, is used to investigate the circumstances under which nontrivial collective behavior occurs . It shows how a reversal in the dominant direction of information flow from bottom-up to top-down is correlated with the emergence of collective behavior of the populations.

{\it FROM INSTRUCTIONS One to two paragraphs that summarize the purpose, methods and analysis of the Walker paper and how you extend that analysis. }

KELLIN:  State HOW DO WE EXTEND THIS ANALYSIS?

@@

\subsection{Sensitive Dependence on Initial Conditions for the Logistic Map}

Depending on the value of the reproductive fitness $r$, the logistic map will generate a population size that is stable, periodic, or chaotic over time.  The behavior of the logistic map can vary a great deal with small changes in $x_0$.  This phenomenon is known as {\it sensitive dependence on initial conditions}, where small changes in the input lead to large changes in the output.  In figure \ref{figure1abcd}, we plot four times series showing population versus time for $50$ time steps.   
%r, initial x values and cross entropy
% a = 0.7522
% b = 
% ab mutual information  0.7522

%cd mutual information   0.0471
The mutual information measures the dependence of one random variable on another random variable.   For two random variables $X$ and $Y$, we define the mutual information to be
\begin{eqnarray}
I(X,Y) = \sum_{y \in Y} \sum_ {x \in X} p(x,y) { \log p(x,y) \over {\log p(x)  \log p(y)} }
\end{eqnarray}
When two random variables are independent, their mutual information is $0$.

We computed the mutual information for the time series in figure \ref{figure1abcd}(a) and figure \ref{figure1abcd}(b).  
The mutual information for the periodic maps in figure \ref{abcd_combined}(a)  is $0.7522$ and the mutual information for the chaotic maps it $0.471$.   This makes sense because after $t=13$, for the periodic maps, you can certainly determine the value one map by knowing the other.  It also makes sense that the mutual information for the chaotic maps should be close to zero because they are chaotic.

To estimate the mutual information, we produced values that could be binned by multiplying each population size by 100 and rounding to the nearest integer.  Then we used the JIDT AutoAnalyzer \cite{JIDT} to compute the mutual information assuming discrete data.  To generate figure \ref{figure1abcd} we used Matlab \cite{Matlab} with logistic map equation from \cite{Mitchell}.

\begin{figure}[H]
\centering
 \includegraphics[width=90mm]{figure1abcd_combined}

  \caption{ Population growth over time showing sensitive dependence on initial conditions.  The mutual information for the periodic map (a) is 0.7522.  The mutual information for the chaotic map (b) is 0.0471.}
 \label{figure1abcd}
\end{figure} 
To test whether binning the data might be adversely affecting our mutual information estimates and since our data was actually real instead of discrete, we also tried IDT?s kernel estimator to measure the mutual information. 
Kernel density estimators are a method for estimating probability densities \cite{Moon}. The JIDT kernel estimator assumes the input data is Gaussian and approximates distributions for all the variables based on this assumption.  One can then use the JIDT AutoAnalyzer to calculate the mutual information between the distributions.   Using the kernel estimator, we estimated the mutual information between the periodic maps to be 1.7086 and the mutual information between the chaotic maps to be 0.7192.   Again, this validates what we see in \ref{figure1abcd} because the chaotic maps have less mutual information than the periodic ones.

\subsection{The Walker Model}
For this project, we used the  model of coupled growth defined in \cite{Walker}. This model uses a logistic growth model on a lattice of populations coupled with influence from the entire population of populations.  For each time step $n$, the size of population $i$ at time step $n$ is defined to be
\begin{eqnarray}
x_{i, n+1} = (1 - \epsilon) f_i(x_{i,n}) + \epsilon m_n, \ \ \ (i=1, 2, \ldots N)
\label{coupling}
\end{eqnarray}
Here $N$ is the number of time steps, $n$ is the current time step (or generation), and $\epsilon$ is the coupling strength of the model to the entire lattice of organisms.   Here the the local growth dynamics of each population $i$ is defined by 
\begin{eqnarray}
f_i(x_{i,n}) = r_i x_{i,n} (1 - {x_{i,n} \over K})
\label{individual}
\end{eqnarray}
Here $r_i$ is the reproductive fitness of population $i$ and $K$ is the carrying capacity.

We model the state of the entire lattice at time $n$ is an average of the states of all the populations.  This is called the instantaneous mean field and defined to be
\begin{eqnarray}
M_n = { 1 \over N} \sum _{j=1}^N x_{j,n}
\label{bigM}
\end{eqnarray}
 The mean-field $m_n$ is a measure of how the individual populations are growing at any time-step $j$.  It is defined to be 
 \begin{eqnarray}
 m_n = {1 \over N} \sum_{j=1}^N f_j(x_{j,n})
 \label{littlem}
\end{eqnarray}
The influence of $m_i$ on the entire system is quantified by the parameter $\epsilon$, the coupling strength.

To do our simulation,  we set the carrying capacity $K$ to be 100 for all populations $i$, the number of populations $N = 1000$ and use $10000$ time steps.   We set the initial population size $x_i, 0$ to be $1$.   We also generated a random fitness $r_i$ for each population $i$.

In figure \ref{prettypicture}, we show the return maps for a selection of coupling strengths $\epsilon$.

% Figure for part 2
 \begin{figure}[t]
 \centering
    \includegraphics[width=100mm]{prettypicture}

    \caption{ Return map for selected values of  the global coupling strength $\epsilon$:  $\epsilon =0$ (magenta), $\epsilon = 0.075$ (red), $\epsilon = 0.1$ (blue), $\epsilon = 0.25$ (orange), $\epsilon = 0.225$ (aqua), $\epsilon = 0.25$ (dark green), $\epsilon = 3 $ (stars) and $\epsilon = 4$ (black)}
 \label{prettypicture}
\end{figure}

To build the return map, equations \ref{coupling},  \ref{individual}, \ref{bigM},  and \ref{littlem} were implemented  in Python using Numpy \cite{numpy} and Matplotlib \cite{matplotlib}.   The return map shows that when the coupling $\epsilon = 0$,  (magenta) there is not much change in how the overall state of the populations changes from time step to time step.  That is, populations  act as isolated populations.   When $\epsilon > 0$, we see a variety of behaviors.

\begin{figure*}[!t]
\centering
\subfloat[]{\includegraphics[width=3.5in]{Fig4a}%
\label{te}}
\hfill
\subfloat[]{\includegraphics[width=3.5in]{Fig4b}%
\label{fig_second_case} }
\caption{(a) Top-down, $T_{M\rightarrow X}$ (solid), and bottom-up, $T_{X\rightarrow M}$ (dashed), causal information transfer as a function of the global coupling parameter $\epsilon$ for coupled logistic maps. (b) The difference (Top-down minus bottom-up) in causal information transfer for $30$ coupled systems as well as the mean as a function of $\epsilon$ (dashed). Vertical line at $\epsilon = 0.3$ illustrates the directional reversal of causal information transfer.}
\label{te_avg}
\end{figure*}

It should be noted that there is quite a bit of dependence on the initial fitness values $r_i$   In different runs of the return map, $\epsilon = 0.25$ (dark green) and and $\epsilon = 0.225$ (aqua) populations sometimes form loops and there can be some scatter instead of two fixed points when $\epsilon = 0.3$.  Overall, we see that coupling strengths $0, 0.075, 0.1$ and $0.4$ produce predictably shaped regions, while there can be variation in the regions produced by $\epsilon = 0.2,  \epsilon = 0.225$ and $\epsilon = 0.3$.

@@

While return maps that look like figure \ref{prettypicture} seemed to occur more often, it would be interesting to look at the distributions of fitness values $r_i$ that produce the more exotic return maps.  This is outside of the scope of this paper.

\subsection{Direction of Information Flow}
\noindent In order to assess that direction of information flow, from Global to Local scales and vice versa, the authors in \cite{Walker} take advantage of {\it Transfer Entropy}. Transfer Entropy is defined mathematically in \cite{Schreiber} as
$$T^{(k)}_{Y \rightarrow X} = \sum_n p(x_{n+1}, x_n^{(k)}, y_n^{(k)}) \log \left[ \frac{p(x_{n+1}|x_n^{(k)}, y_n^{(k)})}{p(x_{n+1}|x_n{(k)}}\right]$$
Where $x_n^{(k)}$ is an embedded state in the $k$-dimensional phase spate defined as $x_n^{(k)} = (x_n, \cdots, x_{n-k+1})$. \\
\begin{figure}[t]
\centering
 \includegraphics[width=90mm]{Fig5}

@@

\noindent Although Transfer Entropy is a fine candidate to estimate directionality of this coupling, recent work has focused on improving estimation methods. In their 2015 paper, Gencaga et. al propose new "statistically robust" methodology to estimate Transfer Entropy \cite{recipe}. Relevant here, is the considerable improvement for continuous processes, where discretizing the data can lead to a distortion in the underlying probability distribution. For this reason, we choose to use the state of the art TransferEntropy package in R and claim based on this work that these estimates of Transfer Entropy are to be preferred. \\
x

\noindent In Figure \ref{te}, we see qualitatively similar results as to what was demonstrated in \cite{Walker}. The quantitative difference in the location and magnitude of the peaks can be attributed to either chance or estimation method. For low values of the global coupling parameter ($\epsilon$), we see information flow is stronger in the bottom-down direction (from local to global scale). As $\epsilon$ increases the direction reverses, and top-down causation begins to dominate. We note again that we are qualitatively in agreement with the previous work \cite{Walker}. There is however a subtle quantitative difference in our results. It appears from Figure \ref{te} that this reversal of causation occurs at a value of $\epsilon = 0.3$. In an effort to analyze this claim more rigorously, we repeat the entire simulation and estimation procedure $30$ times. The right panel of Figure \ref{te_avg} shows the difference $T_{M\rightarrow X} - T_{X \rightarrow M}$ as a function of $\epsilon$ for 30 different simulations. The mean is also plotted, and we can see clearly that this reversal of information flow occurs right at $\epsilon = 0.3$. Finally, we note that this result is intuitively satisfactory since it coincides with the black stars in Figure \ref{prettypicture}, the value of $\epsilon$ for which the system is completely synchronized.

\subsection{Additional Analysis}
\noindent Here we continue our exploration of the relationship between local and global population dynamics. We will use the metric displayed in Figure \ref{te_avg} $T_{M\rightarrow X} - T_{X\rightarrow M}$ to discuss the direction of causal inforamtion transfer between the two scales. We break our discussion into 3 cases.
\begin{itemize}
\item[i)] For small values of the global coupling strength, say $\epsilon < 0.3$, Figure \ref{te_avg} shows that our metric takes negative values. This indicates that causal information is flowing from the local to the global scale. The top row of Figure \ref{timeseries} shows that the individual population is chaotic, but the mean trajectory oscilates roughly between 3 values ($0.4$, $0.6$ and $0.6$). This is what gives the red points their behavior in Figure \ref{prettypicture}.
\item[ii)] For $\epsilon \approx 0.3$, the system synchronizes. In fact, we can compute the correlation between the trajectories in the middle row of Figure \ref{timeseries}, and see that it is nearly equal to one ($0.9879)$. Based on Figure \ref{prettypicture} we sould expect these series to oscillate between precisely two values. This particular run had four attractors (black stars) instead of two.
\item[iii)] For large values of the global coupling strength, the global dynamics begin to dominate. There is still a near perfect correlation between $M$ and $X$, but now even the global dynamics (the mean) behaves chaotically.
\end{itemize}
For small values of the Global Coupling Strength, say $\epsilon < 0.3$, this metric takes negative values, indicating that information is flowing from the local to the global scale. The top row of Figure \ref{timeseires} shows that the individual populations appear chaotic, but the mean trajectory appears periodic. This is similar to the behavior illus

\section{Extra Credit}  Extra credit.

@@

While each of us touched all sections of this paper, Vanessa Job was the primary author of the introduction and the sections about figures 1 and 2, as well as the code that produced the the results in these sections.   Kellin Rumsey was the primary author of the sections that produced figures 3 and 4, as well as the code that produced the results in these sections.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
{\bf CHECK THAT BIBITEMS HAVE CORRECT FORMAT }
\begin{thebibliography}{1}

@@

\bibitem{Walker} Evolutionary Transitions and Top-Down Causation

\bibitem{Mitchell} ]M. Mitchell, Complexity. Oxford: Oxford University Press, 2011, pp. 27-37.

@@

\bibitem{moon}
Moon, Young-Il, Balaji Rajagopalan, and Upmanu Lall.  Estimation of mutual information using kernel density estimators.  {\it Physical Review E}, Volume 52, Number 3, September 1995.

\bibitem{Schreiber} FILL THIS IN

\bibitem{TE} FILL THIS IN

\bibitem{recipe} FILL THIS IN
\end{thebibliography}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
In this paper, we investigate the direction of causal information transfer from global to local scales and vice versa. As in previous work, we use a system of coupled logistic maps as a toy example of population growth. We explore the collective behavior which arises in a variety of forms for different values of a global coupling parameter. We utilize state-of-the-art estimation procedures to infer both top-down and bottom-up transfer entropies. Finally, we analyze these results as they pertain to global coupling, and attempt to uderstand their implications.
\end{abstract}

@@

\noindent For individual populations, we assume birth rate, death rate, and  carrying capacity are all fixed.  Define $k$ to be the carrying capacity of a population and $n_t$ to be size of the population at time $t$. Then the size of the population at generation $t$ is modeled by 
\begin{eqnarray}
n_{t+1} = (\textnormal{birthrate} - \textnormal{deathrate}) (k n_t - n_t^2) / k 
\end{eqnarray}
If we define the population size relative to the carrying capacity as  $x_t = n_t / k$ and define the reproductive fitness  $r = birthrate - deathrate$, then we  get the following equation for $x_t$: 
\begin{eqnarray}
 x_{t+1} = r  x_t (1 - x_t), \ \ \ \ \  t=0, 1, 2, \ldots . 
\label{logistic}
\end{eqnarray}
This is the logistic map as defined in \cite{Mitchell}.  In what follows, we will refer to $x_t$ as the population size, with the understanding that it will always be the ratio of the actual population size to the carrying capacity. \\

\noindent In \cite{Walker}, the authors investigate information flow between higher and lower levels of organization, between individuals of a population and the population considered as a whole. 
They study a model that consists of a population of individual populations and examine the direction of the causal information transfer, where 
"non-trivial collective behavior is associated with the degree to which local elements receive information from the global network." \cite{Walker} \\

\noindent Their model, a lattice of globally coupled logistic maps, is used to investigate the circumstances under which nontrivial collective behavior occurs. Through the use of transfer entropy, they show the relationship between a global coupling parameter and direction of information flow. They also illustrate the relationship between these dynamics and the emergence of complex collective behavior of the populations.  \\

\noindent In this paper, we seek to reproduce these results and to answer some additional questions. We will explore in detail the behavior of the mean-field as a function of global coupling strength and thus by direction of information flow. In addition, we will attempt to improve on the estimation of transfer entropy via the use of kernel density estimators (2A) and Kraskov's estimator (section 2D). Finally, we consider the direction of information as described by transfer entropy in the average case, rather than relying on making inferences from a single simulation.

\noindent

@@

\subsection{Sensitive Dependence on Initial Conditions}
\noindent Depending on the value of the reproductive fitness $r$, the logistic map will generate a population size that is stable, periodic, or chaotic over time.  The behavior of the logistic map can vary a great deal with small changes in $x_0$.  This phenomenon is known as {\it sensitive dependence on initial conditions}, where small changes in the input lead to large changes in the output.  In figure \ref{figure1abcd}, we plot four times series showing population versus time for $50$ time steps.   \\
%r, initial x values and cross entropy
% a = 0.7522
% b = 
% ab mutual information  0.7522

%cd mutual information   0.0471

\noindent The mutual information measures the dependence of one random variable on another random variable.   For two random variables $X$ and $Y$, we define the mutual information to be
\begin{eqnarray}
I(X,Y) = \sum_{y \in Y} \sum_ {x \in X} p(x,y) { \log p(x,y) \over {\log p(x)  \log p(y)} }
\end{eqnarray}
When two random variables are independent, their mutual information is $0$. \\

\noindent We first estimated the mutual information for the time series in figure \ref{figure1abcd}(a) and figure \ref{figure1abcd}(b) using binning.  
The mutual information for the periodic maps in figure \ref{figure1abcd}(a)  is $0.7522$ and the mutual information for the chaotic maps is $0.0471$.   This makes sense because after $t=13$, for the periodic maps, you can certainly determine the value one map by knowing the other.  It also makes sense that the mutual information for the chaotic maps should be close to zero because they are chaotic.  \\

\noindent To estimate the {\it mutual information} with binning, we produced values that could be binned by multiplying each population size by 100 and rounding to the nearest integer.  Then we used the JIDT AutoAnalyzer \cite{JIDT} to compute the mutual information assuming discrete data.  To generate figure \ref{figure1abcd} we used Matlab with the logistic map equation from \cite{Mitchell}.

\begin{figure}[t]
\centering
 \includegraphics[width=90mm]{figure1abcd_combined}
  \caption{ Population growth over time showing sensitive dependence on initial conditions.  The mutual information for the periodic map (a) is 0.7522.  The mutual information for the chaotic map (b) is 0.0471.}
 \label{figure1abcd}
\end{figure} 
\noindent To test whether the binning of the continuous data might be adversely affecting our estimation of mutual information, we also tried JIDT's kernel estimator to estimate the mutual information. In {\it kernel density estimation} (KDE), a Gaussian (bell) curve is centered at each data point. These Gaussian curves are collected, aggregated and finally normalized so that they integrate to one. The result is a smooth continuous estimate of the probability density, given only a finite amount of data\cite{Moon}.  One can then use the JIDT AutoAnalyzer to calculate the mutual information between the distributions.   Under this method, we found the mutual information to be 1.7086 for the periodic maps, and 0.7192 for the chaotic maps.  This reinforces what we see in Figure \ref{figure1abcd} because the chaotic maps have less mutual information than the periodic ones.

\subsection{The Walker Model}
\noindent For this project, we used the toy model of coupled logistic growth defined in \cite{Walker}. This model uses the logistic map on a lattice of populations coupled with influence from the dynamics of the mean-field.  For each time step $n$, the size of population $i$ at time step $n$ is defined to be
\begin{eqnarray}
x_{i, n+1} = (1 - \epsilon) f_i(x_{i,n}) + \epsilon m_n, \ \ \ (i=1, 2, \ldots N)
\label{coupling}
\end{eqnarray}
Here $N$ is the number of time steps, $n$ is the current time step (or generation), and $\epsilon$ is the coupling strength of the model to the entire lattice of organisms.   Here the the local growth dynamics of each population $i$ is defined by 
\begin{eqnarray}
f_i(x_{i,n}) = r_i x_{i,n} \left(1 - {x_{i,n} \over K}\right)
\label{individual}
\end{eqnarray}
Here $r_i$ is the reproductive fitness of population $i$ and $K$ is the carrying capacity.    
% Figure for part 2
 \begin{figure}[t]
 \centering
    \includegraphics[width=100mm]{prettypicture}
    \caption{ Return map for selected values of  the global coupling strength $\epsilon$:  $\epsilon =0$ (magenta), $\epsilon = 0.075$ (red), $\epsilon = 0.1$ (blue), $\epsilon = 0.25$ (orange), $\epsilon = 0.225$ (aqua), $\epsilon = 0.25$ (dark green), $\epsilon = 3 $ (stars) and $\epsilon = 4$ (black)}
 \label{prettypicture}
\end{figure} 
\noindent We model the state of the entire lattice at time $n$ by an average of the states of all the populations.  This is called the {\it instantaneous mean field}, defined to be
\begin{eqnarray}
M_n = { 1 \over N} \sum _{j=1}^N x_{j,n}
\label{bigM}
\end{eqnarray}
 The dynamics of the mean-field, $m_n$, is a measure of how the individual populations are growing at any time-step $n$. 
 \begin{eqnarray}
 m_n = {1 \over N} \sum_{j=1}^N f_j(x_{j,n})
 \label{littlem}
\end{eqnarray}
The influence of $m_i$ on the entire system is quantified by the parameter $\epsilon$, the coupling strength.  \\

\begin{figure*}[!t]
\centering
\subfloat[]{\includegraphics[width=3.5in]{Fig4a}%
\label{te}}
\hfill
\subfloat[]{\includegraphics[width=3.5in]{Fig4b}%
\label{fig_second_case} }
\caption{(a) Top-down, $T_{M\rightarrow X}$ (solid), and bottom-up, $T_{X\rightarrow M}$ (dashed), causal information transfer as a function of the global coupling parameter $\epsilon$ for coupled logistic maps. (b) The difference (Top-down minus bottom-up) in causal information transfer for $30$ coupled systems as well as the mean as a function of $\epsilon$ (dashed). Vertical line at $\epsilon = 0.3$ illustrates the directional reversal of causal information transfer.}
\label{te_avg}
\end{figure*}
\noindent 
For simulation, we fix the carrying capacity to be $K=100$, and we simulate $N=1000$ populations for ten-thousand time steps. For each population, we initialize $x_{i, 0} = 1$ and we generate a fitness value $r_i$ independently and uniformly on the interval $[3.9, 4.0]$. Figure \ref{prettypicture} shows return maps for various values of the global coupling strength $\epsilon$.  \\

\noindent To build the return map, equations \ref{coupling},  \ref{individual}, \ref{bigM},  and \ref{littlem} were implemented  in Python using Numpy \cite{numpy} and Matplotlib \cite{matplotlib}.   The return map shows that when the coupling $\epsilon = 0$,  (magenta) there is not much change in how the overall state of the populations changes from time step to time step.  That is, populations  act independently of eachother.   When $\epsilon > 0$, we see a variety of interesting behaviors.

\noindent It should be noted that the simulation is sensitive to the simulated fitness values $r_i$.  In different runs of the return map, $\epsilon = 0.25$ (dark green) and and $\epsilon = 0.225$ (aqua) populations sometimes form loops. Moreover, there can be more than two attractors when $\epsilon = 0.3$.  Overall, we see that coupling strengths $0, 0.075, 0.1$ and $0.4$ produce consistently reproducible patterns, while there can be variation between runs in the regions produced by $\epsilon = 0.2,  \epsilon = 0.225$ and $\epsilon = 0.3$.

@@

\noindent While return maps that look like figure \ref{prettypicture} seemed to occur more often, it would be interesting to look at the distributions of fitness values $r_i$ that produce the more exotic return maps. This is beyond the scope of this paper.

\subsection{Direction of Information Flow}
\noindent In order to assess the direction of information flow from global to local scales and vice versa, the authors in \cite{Walker} take advantage of {\it transfer entropy}. Transfer entropy is defined mathematically in \cite{Schreiber} as
$$T^{(k)}_{Y \rightarrow X} = \sum_n p(x_{n+1}, x_n^{(k)}, y_n^{(k)}) \log \left[ \frac{p(x_{n+1}|x_n^{(k)}, y_n^{(k)})}{p(x_{n+1}|x_n{(k)}}\right]$$
Where $x_n^{(k)}$ is an embedded state in the $k$-dimensional phase space defined as $x_n^{(k)} = (x_n, \cdots, x_{n-k+1})$. \\
\begin{figure}[t]
\centering
 \includegraphics[width=90mm]{Fig5}

@@

\noindent Although transfer entropy is a fine candidate for estimating the directionality of this coupling, recent work has focused on improving estimation methods. In their 2015 paper, Gencaga et. al propose new "statistically robust" methodology to estimate Transfer Entropy \cite{recipe}. Relevant here is the considerable improvement for continuous processes, where discretizing the data can lead to a distortion in the underlying probability distribution. For this reason, we choose to use the state-of-the-art "Kraskov Estimtor", supported by the TransferEntropy\cite{TE} package in R, and claim based on this work that these estimates of transfer entropy are to be preferred for this problem. \\

\noindent In Figure \ref{te}, we see qualitatively similar results as to what was demonstrated in \cite{Walker}. The quantitative difference in the location and magnitude of the peaks can be attributed to either chance or estimation method. For low values of the global coupling parameter ($\epsilon$), we see information flow is stronger in the bottom-down direction (from local to global scale). As $\epsilon$ increases, the direction reverses and top-down causation begins to dominate. We note again that we are qualitatively in agreement with the previous work \cite{Walker}. There is however a subtle quantitative difference in our results. It appears from Figure \ref{te} that this reversal of causation occurs at a value of $\epsilon = 0.3$. In an effort to analyze this claim more rigorously, we repeat the entire simulation and estimation procedure $30$ times. The right panel of Figure \ref{te_avg} shows the difference $T_{M\rightarrow X} - T_{X \rightarrow M}$ as a function of $\epsilon$ for 30 different simulations. The mean is also plotted, and we can see clearly that this reversal of information flow occurs right at $\epsilon = 0.3$. Finally, we note that this result is intuitively satisfactory since it coincides with the black stars in Figure \ref{prettypicture}, the value of $\epsilon$ for which the system is completely synchronized.

\subsection{Additional Analysis}
\noindent Here we continue our exploration of the relationship between local and global population dynamics. We will use the metric displayed in Figure \ref{te_avg}, $T_{M\rightarrow X} - T_{X\rightarrow M}$, to discuss the direction of causal inforamtion transfer between the two scales. We break our discussion into 3 cases.
\begin{itemize}
\item[i)] For small values of the global coupling strength, say $\epsilon < 0.3$, Figure \ref{te_avg} shows that our metric takes negative values. This indicates that causal information is flowing from the local to the global scale. The top row of Figure \ref{timeseries} shows that the individual population is chaotic, but the mean trajectory oscilates roughly between 3 values ($0.4$, $0.6$ and $0.6$). This is what gives the red points their behavior in Figure \ref{prettypicture}.
\item[ii)] For $\epsilon \approx 0.3$, the system synchronizes. In fact, we can compute the correlation between the trajectories in the middle row of Figure \ref{timeseries}, and see that it is nearly equal to one ($0.9879)$. Based on Figure \ref{prettypicture} we sould expect these series to oscillate between precisely two values. This particular run had four attractors (black stars) instead of two.
\item[iii)] For large values of the global coupling strength, the global dynamics begin to dominate. There is still a near perfect correlation between $M$ and $X$, but now even the global dynamics (the mean) behaves chaotically.
\end{itemize}

\noindent The mean-field only displays true near-periodic behavior when $\epsilon = 0.3$, when it oscilates between a small number of attractors. Although the mean population will never be truly asymptotic, it can be "pseudo-asymptotic". For instance when $\epsilon = 0$ and when $\epsilon = 0.1$, we see from Figure \ref{prettypicture} that the mean-field bounces around a single attractor. For other values of $\epsilon$, $\epsilon = 0.25$ and $\epsilon = 0.075$ in particular, the mean field bounces around in a "pseudo-periodic" fashion. When $\epsilon > 0.4$, we see that the mean-field is behaving chaotically itself.  \\

\noindent We finish this section with a final discussion of the $\epsilon = 0.075$. In our example of Figure \ref{timeseries}, we see that the individuals appear to be behaving nearly-chaotically, while the global trend oscilates roughly between $3$ attracting regions. We argue that it is unlikely that top-down information transfer is driving this behavior. This is also coherent with our conclusions based on Figure \ref{te_avg}, where we see that bottom-up information transfer is stronger for this value of $\epsilon$.

%\section{Extra Credit}  Extra credit.

@@

\noindent Each member of the team had an influence over all sections of this paper Precisely, Vanessa Job was the primary author of the introduction and the sections about figures 1 and 2, as well as the code that produced the the results in these sections.   Kellin Rumsey was the primary author of the sections that produced figures 3 and 4, as well as the code that produced the results in these sections.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{Walker} Walker SI, Cisneros L, Davies PCW. 2012 Evolutionary transitions and top-down causation. {\it Proc. Artif Life} {\bf XIII}, 283-290.

@@

\bibitem{Mitchell} M. Mitchell, Complexity. Oxford: Oxford University Press, 2011, pp. 27-37.

\bibitem{Moon}
Moon, Young-Il, Balaji Rajagopalan, and Upmanu Lall.  Estimation of mutual information using kernel density estimators.  {\it Physical Review E}, Volume 52, Number 3, September 1995.

@@

\bibitem{Schreiber} Schreiber, T. (2007) Meausring information transfer. {\it Phys. Rev. Lett.}, 85:461

\bibitem{recipe} D. Gencaga, K. H. Knuth, and W. B. Rossow, A recipe for the estimation of information flow in a dynamical system {\it Entropy} 17, 438 (2015)

\bibitem{TE} G. H. Torbati, G. Lawyer, TransferEntropy: The Transfer Entropy Package. R package, Version 1.4

\end{thebibliography}
